---
title: "Exploring molecular adducts in MSI with mass2adduct"
author: "Brandon Seah"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mass2adduct}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Background: Molecular adducts in mass spectrometry imaging

In mass spectrometry imaging (MSI), adducts can form between target molecules (e.g. metabolites) and other substances such as matrix or salt ions.

This package presents tools for counting and identifying possible adducts in MS data, and accompanies Janda et al. (in prep.).

## Import data

Raw MSI data come in different formats, including the emerging standard imzML format, which is based on XML. For the purposes of mass2adduct, we are interested in data after preprocessing (e.g. baseline correction, peak-binning/picking). This should be exported as plain-text CSV files from standard MSI software such as SCiLS or MSiReader.

You can find an example CSV file in the folder `inst/extdata` in the source package. The columns represent mass peaks, with m/z values as column names, and rows represent pixels, with row names (these can be arbitrary as they are not used in this workflow). The entries in the table represent intensity values for a given peak and pixel.

```{r}
library(mass2adduct)
d <- msimat(system.file("extdata","msi.csv",package="mass2adduct"),sep=";")
```

If the data matrix is very large, it may need to be reformatted to be loaded into memory during an R session.

Typing the object name or using the `summary()` function gives you a summary of its contents.

```{r}
d # Object of class massdiff
```

## Tabulate all pairwise mass differences

Take all possible pairs of masses and calculate the mass difference for each pair. These mass differences represent potential molecular adducts.

```{r}
d.diff <- massdiff(d) # Returns object of classes data.frame and massdiff
```

View a summary of the contents:

```{r}
summary(d.diff)
```

Preview the first few lines of the results:

```{r}
head(d.diff)
```

There are three columns:

 * `A` is the parent ion peak mass
 * `B` is the putative adduct ion peak mass (higher than A)
 * `diff` is the difference between A and B

At this point, the intensity information is not yet used.

## Quickly identify possible adducts

Of course, most pairs do not represent real or meaningful chemical transformations. However, we expect that common chemical transformations, e.g. the difference between carbon-12 and carbon-13 isotopes in organic compounds, will be more often encountered than the background "noise". 

### Visualize with a histogram

A histogram of all mass difference values will provide an overview. We expect that interesting transformations or adducts will be peaks in the histogram, i.e. they are more commonly encountered as parent-adduct ion pairs than random values.

```{r, fig.width=7, fig.height=4}
d.diff.hist <- hist(d.diff) # Object of classes histogram and massdiffhist
plot(d.diff.hist,labels=10)
```

The plot shows peaks representing common chemical transformations in this sample. The built-in data set `adducts` contains a list of biologically-relevant chemical species that might occur in biological samples. With the `labels=10` option, you can label the top 10 peaks which have close matches to that list.

We can examine certain areas of the histogram in more detail.

```{r, fig.width=7, fig.height=4}
plot(d.diff.hist, labels=10, xlim=c(100,200))
```

Also note that the background of likely non-meaningful mass differences is not uniformly distributed, but instead has peaks at integer values. This can be understood from the fact that most chemical masses are themselves close to integer values, hence the differences between them should be too. 

### Summarize possible matches to chemical adducts

The package comes with a small data set of common molecular adducts, called adducts (use the command help(adducts) to see a description). Users can supply their own custom sets of adducts as long as they are in the same format (as a data.frame with three columns: name, formula, mass)

The following function looks for known adducts by finding the closest-matching bin in the mass difference histogram produced above. It reports the number of counts (i.e. how many pairs of MS peaks have that mass difference) and the quantile.

Note that quantile values will usually be quite high because the majority of mass differences have zero to few counts.

```{r}
head(adductMatch(d.diff.hist),n=10) # Show the top ten matches to known adducts
```

However this only shows known adducts. What about peaks in the histogram which don't have a good match to the reference list? `topAdducts` ranks mass differences by their occurrences, and reports them in descending order, as well as matches to known adducts, if any

```{r}
topAdducts(d.diff.hist, n=10) # Report top ten hits
```

## Examine specific adducts in detail

Using the histogram gives us a useful overview and helps us to see if there are possible chemical transformations that we do not know about (i.e. unlabelled peaks). However, there are some drawbacks. The histogram bins massdiff values into bins of fixed width, whereas we know that the mass resolution of a mass spectrometer varies with the mass. What do we do if we would like to identify massdiffs that correspond to a specific adduct of interest?

## Test for correlations between parent and adduct ions


